The five axes of the Turtle: symmetry and asymmetry in NGC~6210

### 2020-06-18 ms.tex
Nearly finish changing abstract

\title{The complex structure and peculiar internal motions of the planetary nebula NGC 6210}

\author[Henney et al.]{
  William J. Henney,\(^1\)\thanks{
    w.henney@irya.unam.mx,
    jal@astro.unam.mx,
    tere@astro.unam.mx,
    richer@astro.unam.mx
  }
  J. A. López\(^2\)\footnotemark[1]
  Ma.\ T. García-Díaz,\(^2\)\footnotemark[1]
  and M. G. Richer\(^2\)\footnotemark[1]
  \\
  \(^1\)\foreignlanguage{spanish}{
    Instituto de Radioastronomía y
    Astrofísica, Universidad Nacional Autónoma de México, Apartado
    Postal 3-72, 58090 Morelia, Michaoacán, Mexico}
  \\
  \(^2\)\foreignlanguage{spanish}{
    Instituto de Astronomía,
    Universidad Nacional Autónoma de México,
    Ensenada, Baja California, 22800, México}
}
% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

@@

\begin{abstract}
  A comprehensive kinematic study has been carried out on the planetary nebula (PN) NGC 6210. Multiple long-slits, echelle spectra have been obtained over the face of this nebula mapping its full shell structure, the opposite pairs of extended, twisted arms and the bipolar collimated outflows and bullets. Public {\it HST} imagery has been used to identify kinematic elements with structural components. The long-slit spectroscopic information has been combined into channel maps that greatly facilitate visualizing the otherwise intricate expansion pattern of this  planetary nebula. The global morphology of NGC 6210 is  reminiscent of other planetary nebulae with extended X-ray emission, suggesting the presence of a central hot bubble produced by shocked stellar wind. In spite of the dramatic structure of this PN substantial expanding radial motions are only found in the material surrounding the central, inner shell. The relatively slow radial motions of the collimated outflows and the bullet-like knots indicate that they are  moving away from the core very close to the plane of the sky, nearly perpendicular to the line of sight. The direction and ages of the various components, derived from their expanding velocities indicate that NGC 6210 has been strongly influenced by the  action of a bipolar, rotating, episodic outflow.
\end{abstract}

@@

In order to investigate the kinematics of the nebula in detail,
we have measured the velocities of distinct emission components in each slit spectrum
and organized them into broad systems based on their location, morphology and degree of ionization.

@@

The final axis, E, is the ejection axis of the low-ionization knot complexes,
which dominate the appearance of the nebula in the \nii{} line.
Axis~E is highly inclined (\(\approx\ang{70}\)) to the plane of the sky
and the low-ionization outflow is less well collimated than the lobes.
These combine to make the axis hard to identify in images,
although it is clearly revealed in the longslit spectra.
Furthermore, while A, B, C, and D are symmetry axes,
axis~E would be better described as an \emph{asymmetry} axis,
since the kinematic timescales are utterly different between the redshifted and blueshifted sides.
The red semi-axis (predominantly N and W of the central star)
was active with outflow velocities of \SI{30}{km.s^{-1}}
from \num{8000} to \SI{4000}{yr} ago,
that is, before the planetary nebula phase.
In contrast, the blue semi-axis (predominantly S and E)
was active with outflow velocities of \SI{50}{km.s^{-1}}
from \num{1200} to \SI{2700}{yr} ago,
making it the kinematically youngest component of the nebula.
The density of the low-ionization knots is too high for them to be explained by shocked ionized halo material,
so we propose that they represent a dense massive outflow from the central star system.
Their total mass is \SI{0.06}{\msun}, or about 30\% of the mass of the nebular shells.

NGC~6210 is the product of a relatively low-mass \(\approx \SI{1.25}{\msun}\) progenitor
and its post-AGB central star is halfway along its high-luminosity heating track.
In some ways it is very similar to much simpler nebulae such as NGC~1535 and IC~3568,
but its rich variety of bipolar and monopolar outflows,
combined with marked asymmetries in the nebular shells,
set it apart from most other nebulae.
An interacting triple star system is a natural explanation for the Turtle's many peculiarities,
but the exact mechanism is unknown
and direct evidence for multiplicity of the central star is still lacking.
We hope that our detailed elucidation of the Turtle's complicated history of anisotropic mass loss will provide inspiration for future modeling efforts.

>>>>

\title{
  The five axes of the Turtle: symmetry and asymmetry in NGC~6210
  % The complex structure and peculiar internal motions of the planetary nebula NGC 6210
}

\author[Henney et al.]{
  William J. Henney,\(^1\)\thanks{
    w.henney@irya.unam.mx,
    jal@astro.unam.mx,
    tere@astro.unam.mx,
    richer@astro.unam.mx
  }
  J. A. López,\(^2\)\footnotemark[1]
  Ma.\ T. García-Díaz,\(^2\)\footnotemark[1]
  and M. G. Richer\(^2\)\footnotemark[1]
  \\
  \(^1\)\foreignlanguage{spanish}{
    Instituto de Radioastronomía y
    Astrofísica, Universidad Nacional Autónoma de México, Apartado
    Postal 3-72, 58090 Morelia, Michaoacán, Mexico}
  \\
  \(^2\)\foreignlanguage{spanish}{
    Instituto de Astronomía,
    Universidad Nacional Autónoma de México,
    Ensenada, Baja California, 22800, México}
}
% These dates will be filled out by the publisher
\date{Accepted XXX. Received YYY; in original form ZZZ}

@@

\begin{abstract}
  We carry out a comprehensive kinematic and morphological study
  of the spectacularly asymmetrical planetary nebula:
  NGC~6210, known as the Turtle Nebula.
  We study the three-dimensional structure and kinematics of the nebula's shells, lobes, knots, and haloes
  by combining radial velocity mapping from multiple long-slit spectra
  with proper motion measurements from multi-epoch imaging.
  We find that the nebula has five separate ejection axes, which we label A~to~E.\@
  Axes~A and B are major axes of the bipolar wind-blown inner shell
  and the lop-sided, elliptical, fainter, but more massive intermediate shell, respectively.
  Axes~C and D (and also A) are bipolar flow axes that form the point symmetric,
  high-ionization outer lobes, all with inclinations close to the plane of the sky.
  Axis~E, which is inclined close to the line of sight,
  is the axis of a loosely collimated outflow of low-ionization knots.
  % A comprehensive kinematic study has been carried out on the planetary nebula (PN) NGC 6210. Multiple long-slits, echelle spectra have been obtained over the face of this nebula mapping its full shell structure, the opposite pairs of extended, twisted arms and the bipolar collimated outflows and bullets. Public {\it HST} imagery has been used to identify kinematic elements with structural components. The long-slit spectroscopic information has been combined into channel maps that greatly facilitate visualizing the otherwise intricate expansion pattern of this  planetary nebula. The global morphology of NGC 6210 is  reminiscent of other planetary nebulae with extended X-ray emission, suggesting the presence of a central hot bubble produced by shocked stellar wind. In spite of the dramatic structure of this PN substantial expanding radial motions are only found in the material surrounding the central, inner shell. The relatively slow radial motions of the collimated outflows and the bullet-like knots indicate that they are  moving away from the core very close to the plane of the sky, nearly perpendicular to the line of sight. The direction and ages of the various components, derived from their expanding velocities indicate that NGC 6210 has been strongly influenced by the  action of a bipolar, rotating, episodic outflow.
\end{abstract}

@@

In order to investigate the kinematics of the nebula in detail,
we have measured the velocities of distinct emission components in each slit spectrum
and organized them into broad systems based on their location, morphology and degree of ionization.
Maps of the individual slit positions,
marked with the position and velocity of each component,
are presented in Appendix~\ref{sec:veloc-feat-indiv}.

@@

The final axis, E, is the ejection axis of the low-ionization knot complexes,
which dominate the appearance of the nebula in the \nii{} line.
Axis~E is highly inclined (\(\approx\ang{70}\)) to the plane of the sky
and the low-ionization outflow is less well collimated than the lobes.
These combine to make the axis hard to identify in images,
although it is clearly revealed in the longslit spectra.
Furthermore, while A, B, C, and D are symmetry axes,
axis~E would be better described as an \emph{asymmetry} axis,
since the kinematic timescales are utterly different between the redshifted and blueshifted sides.
The red semi-axis (predominantly N and W of the central star)
was active with outflow velocities of \SI{30}{km.s^{-1}}
from \num{8000} to \SI{4000}{yr} ago,
that is, before the planetary nebula phase.
In contrast, the blue semi-axis (predominantly S and E)
was active with outflow velocities of \SI{50}{km.s^{-1}}
from \num{1200} to \SI{2700}{yr} ago,
making it the kinematically youngest component of the nebula.
The density of the low-ionization knots is too high for them to be explained by shocked ionized halo material,
so we propose that they represent dense bullets ejected from the central star system.
Their total mass is \SI{0.06}{\msun}, or about 30\% of the mass of the nebular shells.

NGC~6210 is the product of a relatively low-mass (\(\approx \SI{1.25}{\msun}\)) progenitor
and its post-AGB central star is halfway along its high-luminosity heating track.
In some ways it is very similar to much simpler nebulae such as NGC~1535 and IC~3568,
but its rich variety of bipolar and monopolar outflows,
combined with marked asymmetries in the nebular shells,
set it apart from most other nebulae.
An interacting triple star system is a natural explanation for the Turtle's many peculiarities,
but the exact mechanism is unknown
and direct evidence for multiplicity of the central star is still lacking.
We hope that our detailed elucidation of the Turtle's complicated history of anisotropic mass loss will provide inspiration for future modeling efforts.

\section*{Acknowledgements}
We are grateful for financial support provided by Dirección General de
Asuntos del Personal Académico, Universidad Nacional Autónoma de
México, through grants Programa de Apoyo a Proyectos de Investigación
e Inovación Tecnológica  IN107019 and \textbf{Ensenada grant?}.
